---
title: "Loading Multiple Files from Various Folders in R"
author: "Ed Orlando"
date: 2020-05-14
categories: ["R"]
tags: ["R Markdown", "tidyverse", "purrr", "readxl", "Spotify", "Exportify", "R Studio", "map", "function"]
description: "This blog walks viewers through how to programatically load multiple files from multiple folders using R.  It also shows them how to add the file and folder names to the tibble using the map function."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

### Author: Ed Orlando

### Problem Statement

This tutorial shows you how to load multiple files from multiple folders using R.  As a bonus, this article will also show you how to add the folder and file names to the tibble.

### Load Libraries
Tidyverse and readxl are the only libraries required for this tutorial.

```{r, echo=TRUE, results="hide", warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
```

<br>

### Data Description
In this example, Spotify playlist data was downloaded using [Exportify](<https://rawgit.com/watsonbox/exportify/master/exportify.html>).  Each .xlsx file contains information about the playlist including the artist name, track name, album name, and song duration.  

A sample of one of the the data files is listed below.

```{r, echo=FALSE, results="hide", warning=FALSE, error=FALSE, message=FALSE}
data_sample <- readxl::read_excel(path = "Data_Sources/2020_05_14_Loading_Mult_Files/Genres/Country/90s_country.xlsx")
```
```{r}
data_sample %>% glimpse()
```

<br>

### Folder Structure

Each playlist was previously converted to .xlsx files and placed into the appropriate genre's folder within the R project.  The three genre folders include Country, Hip-Hop, and Rock.

```{r}
list.dirs(path = "Data_Sources/2020_05_14_Loading_Mult_Files/Genres/", 
          full.names = FALSE, 
          recursive  = FALSE)
```

<br>

### File Structure

Each genre folder includes multiple playlist files.  The example below shows the three (3) playlists that exist inside the **Country genre folder**.

```{r}
list.files(path = "Data_Sources/2020_05_14_Loading_Mult_Files/Genres/Country/", 
          full.names = FALSE, 
          recursive  = FALSE)
```

<br>

### Create Function to List All Files in Each Folder

A custom function called **list_files_in_folder()** was created that lists the .xlsx files in each genre folder. The function utilizes the **base::list.files()** function that lists each file under a particular folder.

```{r, echo=TRUE}
list_files_in_folder <- function(genre) {
    
    list.files(path = str_c("Data_Sources/2020_05_14_Loading_Mult_Files/Genres/", genre),
                          pattern = "*.xlsx", 
                          full.names = TRUE) %>% 
    tibble::enframe()
}
```

<br>


### View Files in Country Genre Folder
For example, if we want to view all the files in the Country genre folder, we can now easily pass **"Country"** into the function.

```{r}
list_files_in_folder("Country")
```

<br>

### Create Genre List
Next, the **genre_list()** is generated to includes all three genres.  This list will allow us to map the **list_files_in_folder()** function to multiple genre folders later on.
```{r}
genre_list <- c("Country", "Hip_Hop", "Rock")
```

<br>

### View All Files In All Genre Folders
Using the **purrr::map_df()** function, we can pass the **genre_list()** into the **list_files_in_folder()** function.  The final result produces a tibble with all .xlsx files from all three genre folders we specified above.

```{r}
playlist_tbl <- genre_list %>% 
  purrr::map_df(list_files_in_folder)
```

The **playlist_tbl** below now includes 9 observations.  Each of the three genre folders includes 3 playlist .xlsx files.  

```{r}
playlist_tbl
```

<br>

### Extract "value" Column From playlist_tbl
The value column was extracted from the **playlist_tbl** above and converted into a list.  

```{r}
playlist_list <- playlist_tbl %>%
  select(value) %>%
  as.list()

playlist_list$value
```

<br>

### Read All Excel Files and Add Folder and File Names

The **playlist_list$value** was then mapped into the function below which loaded each .xlsx file.  The code below also added the file name and entire path into a column called **filename**.

```{r results="hide", warning=FALSE, error=FALSE, message=FALSE}
song_tbl <- playlist_list$value %>% 
    map_df(function(add_file_name)
        read_excel(add_file_name) %>%
            mutate(playlist_name = gsub(".xlsx", 
                                        "", 
                                        basename(add_file_name))) %>%
            mutate(filename = add_file_name)
    )
```

```{r}
song_tbl %>% 
  glimpse()
```

<br>

### Separate "filename" Column

Since the entire path and filename are all listed in the **filename** column, the **separate()** function was used to split out three (3) folders, Genre, and the Playlist into five (5) separate columns.

```{r}
song_tidy_tbl <- song_tbl %>% 
            separate(col    = filename, 
                     into   = c("Folder_01", 
                                "Folder_02", 
                                "Folder_03", 
                                "Genre", 
                                "Playlist"),
                     sep    = "/",
                     remove = FALSE)
```

<br>

The final tibble is complete and **tidy** with all **794 songs** listed.

```{r}
song_tidy_tbl %>% glimpse()
```

